The CrypTool Script: \\ Mathematics and Cryptography \\ \vspace{30pt} {\large Background reading\\*[-8pt] for the free elearnig program CrypTool (version 1.3.10)} \vspace{60pt}

### 2005-07-04 script/en/chapters/primes.tex
Update - Bernhard Esslinger June 2005

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O :  P R I M E S
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\hypertarget{Kapitel_2}{}

\section{Prime Numbers}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, June 2003)

@@

% --------------------------------------------------------------------------
\subsection{The search for extremely large primes}
\label{search_for_very_big_primes}   % chap. 2.4

The largest prime numbers known today have several millions digits (M-39), 
which is too big for us to imagine. The number of elementary particles in
the universe is ``only'' estimated to be a $80$-digit number 
\hyperlink{grosord}{(See: overview under \ref{s:grosord}
of various orders of magnitude / dimensions)}.

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers} 
\index{Mersenne!number}
\label{zahlentyp_mersenne}

Almost all known huge prime numbers are special candidates, called
\index{Mersenne Marin} {\em Mersenne numbers}\footnote{%
Marin Mersenne, French priest and mathematician, Sep 08, 1588 $-$ Sep 01, 1648.
\index{Mersenne Marin}
}
of the form $2^p -1,$ where $p$ is
a prime. Not all Mersenne numbers are prime:

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{Factorisation} (expression as a product of prime factors)\footnote{%
Using CrypTool\index{CrypTool} v1.3 you can factorize numbers in the 
following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Demonstration 
\textbackslash{} Factorisation of a Number}.
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221;
 ~ 3,021,377; ~ 6,972,593; \\
13,466,917.
%13,466,917; ~ xxx,xxx,xxx.
\end{array}
$$
Thus
$39$
%$40$
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 38 Mersenne prime numbers we know that this list is complete.
The exponents until the 39th Mersenne prime number have not yet been checked
completely (see chapter \ref{primality_tests}, 
prime number tests\index{Prime number!test}).

The $19$th number with the exponent $4,253$ was the first with at least $1,000$ digits in  decimal system
(the mathematician Samual \index{Yates Samual} Yates coined the expression {\em
titanic} \index{Prime number!titanic} prime for this; it was discovered by
Hurwitz in 1961); the $27$th number with the exponent $44,497$ was the first
with at least $10,000$ digits in the decimal system (Yates coined the expression
\index{Prime number!gigantic}  {\em gigantic} prime for this. These names are
now long outdated).

These numbers can be found at the following URLs:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://reality.sgi.com/chongo/prime/prime_press.html}
               {\texttt{http://reality.sgi.com/chongo/prime/prime\_press.html}}\\
          \href{http://www.utm.edu/}
               {\texttt{http://www.utm.edu/}}
\end{itemize}

\vskip +15 pt
\paragraph{M-37 -- January 1998}\index{Mersenne!prime number!M-37}\mbox{}
\vskip +10 pt

@@

\vskip +15 pt
\paragraph{M-38 -- June 1999}\index{Mersenne!prime number!M-38}\mbox{}
\vskip +10 pt

@@

Right now (Feb. 2003) all exponents smaller than $ 6,972,593 $ have been
tested and double-checked (see home page of the GIMPS project:
{\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}): so we can
be certain, that this is really the 38th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes.

\vskip +15 pt
\paragraph{M13466917 -- December 2001 -- M-39 ?}%
\index{Mersenne!prime number!M-39}\mbox{}
\vskip +10 pt

This number was discovered as 39th Mersenne prime (and already called M-39,
despite it has not been proven yet, whether no further Mersenne prime
numbers between M-38 und M13466917 do indeed exist), $$2^{13,466,917}-1,$$
at December 6, 2001 -- more exactly, the verification of this number,
found at November 14, 2001 by the Canadian student Michael Cameron, was
successfully completed. This number has about 4 million decimal digits
(exactly 4,053,946 digits).
Trying only to print this number 
$$(924947738006701322247758 \cdots 1130073855470256259071)$$
would require around 200 pages in the Financial Times.

\vskip +15 pt
\paragraph{40th Mersenne prime not found -- June 2003}%
\index{Mersenne!prime number!M-40}\mbox{}
\vskip +10 pt

On June 1st a possible Mersenne prime was reported to the GIMPS server,
which was checked afterwards as usual, before it was to be published. 
Unfortunately mid June the initiator and GIMPS project leader George Woltman
had to tell, that two independent verification runs proved the number 
was composite. This was the first false positive report in 7 years.

@@

\vskip +15 pt
\paragraph{GIMPS}\index{GIMPS}\mbox{}
\vskip +10 pt

Discovering the 
39th 
%40th 
Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search)\index{GIMPS} founded in 1996 already discovered for the 
5th 
%6th
time the greatest Mersenne number to be proven being prime
({\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}).

Now more than 130,000 volunteers, amateurs and experts, are working for the GIMPS project. They connect their computers into the so called ``primenet'', organized by the company entropia, to find such numbers using distributed computer programs.

Currently
%(June 2003)
the 
4
%5
largest known primes are Mersenne prime numbers. Only the 
5th 
%6th
biggest known prime is another number type: it belongs to the so called  \hyperlink{generalizedFermatprimes}{generalized Fermat primes}.

% --------------------------------------------------------------------------
\vskip +15 pt
\subsubsection{EFF}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an unknown
donator. The participants are rewarded with a total of 500,000 USD if they find
the longest prime number. In promoting this project, the unknown donator is not
looking for the quickest computer, but rather wants to draw people's attention
to the opportunities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}

@@

Edouard Lucas\index{Lucas Edouard} (1842-1891) held the record for the
longest prime number for over 70 years by proving that $2^{127}-1$ is prime.
No new record is likely to last that long.

% --------------------------------------------------------------------------
\subsection{Prime number tests}
\label{primality_tests}   % chap. 2.5
\index{Prime number!test}

@@

Up to now we have looked for the prime factors in order to decide whether
a number is prime. However, even if the smallest prime factor is enormous, the
search takes too long. Factorising numbers using systematic computational
division or using the \hyperlink{SieveEratosthenes01}{sieve of Eratosthenes} 
\index{Eratosthenes!sieve} is only feasible using current computers for
numbers with up to around $20$ digits in the decimal system.
The biggest number factorized into its 2 almost equal prime factors 
has 158 digits (see \hyperlink{C158-chap3}{chapter \ref{NoteFactorisation}}).

@@

\hypertarget{FermatNumbers01}{}\label{FermatNumbers01}%
In the 17th century, Fermat\footnote{%
Pierre de Fermat, French mathematician, Aug 17, 1601 -- Jan 12, 1665.
\index{Fermat Pierre}
}
\index{Fermat Pierre} wrote to Mersenne \index{Mersenne Marin} that he 
presumed that all numbers of the form 
$$ f(n) = 2^{2^n} + 1 $$ 
are prime for all whole numbers $ n \geq 0$ 
(\hyperlink{FermatNumbers02}{see below}, chapter~\ref{L-FermatNumbers02}).
\index{Number!Fermat}\index{Fermat!number}

@@

\vskip +15 pt
\paragraph{Pseudo prime numbers}%
\index{Prime number!pseudo prime} \index{Number!pseudo prime}%
\hypertarget{HT-Pseudoprimenumber01}{}\label{L-Pseudoprimenumber01}%
\mbox{}
\vskip +10 pt
Numbers n that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$
but are not prime are called {\em pseudo prime numbers}
(i.e. the exponent is not a prime). 
The first pseudo prime number is $$ 341 = 11 \cdot 31 .$$

\vskip +15 pt
\paragraph{Carmichael numbers}%
\index{Number!Carmichael}%
\hypertarget{HT-Carmichael-number01}{}\label{L-Carmichael-number01}%
\mbox{}
\vskip +10 pt
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called {\em Carmichael numbers}. The first of these is
$$ 561 = 3 \cdot 11 \cdot 17 .$$

\vskip +15 pt
\paragraph{Strong pseudo prime numbers}%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller Gary L.}\index{Rabin Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: it is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +20 pt
\subsection{Overview special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

The following enumeration contains the most common ideas for 
``prime number formulae'' and what the status is for big special number types,
trying to either prove primality or determine the prime factors:

% --------------------------------------------------------------------------
\subsubsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 2,000,000 digits are 
    known (M-38\index{Mersenne!prime number!M-38}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}

% --------------------------------------------------------------------------
\subsubsection
   [Generalized Mersenne numbers $f(k,n) = k \cdot 2^n \pm 1$]
   {Generalized Mersenne numbers
   $f(k,n) = k \cdot 2^n \pm 1 $  for $ n $ prime and $ k $ small prime}
   For this first generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}
   there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can
   be performed in practice using software such as the Proths software from
   Yves Gallot\index{Gallot Yves}
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

% --------------------------------------------------------------------------
\subsubsection
   [Generalized Mersenne numbers $ f(b,n) = b^n \pm 1 $ / Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}

@@

    The Fermat number
    $$ f(2,145,351) = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    This prime p is the largest known prime generalized Mersenne 
    number\index{Mersenne!number!generalized} and the
    5th largest known prime number today. 
    It was discovered only a few days after Michael Angel has come across the 
    largest generalized Fermat number (\hyperlink{generalizedFermatprimes}
    {see below}).  This moved GIMPS' \index{GIMPS} first prime discovery,
    M-35 = M1398269, into 7th place.

    This work was done using NewPGen from Paul Jobling's, PRP from 
    George Woltman's, Proth from Yves Gallot's programs\index{Gallot Yves} and
    also the Proth-Gallot group at St. Patrick's College, Dublin.

@@

\vskip +10 pt
% --------------------------------------------------------------------------
\subsection{Density and distribution of the primes}

@@

\item Is there an infinite number of Mersenne prime numbers?\index{Prime number!Mersenne}\index{Mersenne!prime number}

>>>>

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T H R E E :  P R I M E S
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section{Prime Numbers}
\hypertarget{Kapitel_2}{}
\label{Label_Kapitel_2}
(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, June 2003, May 2005)

@@

% --------------------------------------------------------------------------
% --------------------------------------------------------------------------
\vskip + 20pt
\subsection{The search for extremely large primes}
\label{search_for_very_big_primes}   % chap. 3.4

The largest prime numbers known today have several millions digits, 
which is too big for us to imagine. The number of elementary particles in
the universe is estimated to be ``only'' a $80$-digit number 
\hyperlink{grosord}{(See: overview under \ref{s:grosord}
of various orders of magnitude / dimensions)}.

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsubsection{The 10 largest known primes (as of May 2005)} 
\index{Prime number!records}
\label{RecordPrimes}

The following table contains the "top ten" record primes and
a description of its particular number type\footnote{%
An up-to-date version can be found in the internet at
     \href{http://primes.utm.edu/largest.html}
  {\texttt{http://primes.utm.edu/largest.html}}.
}:

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline
	& {\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Description} \\
\hline
	1  & $2^{25,964,951}-1$ & 7,816,230 & 2005 & Mersenne, 42th known \\
	2  & $2^{24,036,583}-1$ & 7,235,733 & 2004 & Mersenne, 41th known \\
	3  & $2^{20,996,011}-1$ & 6,320,430 & 2003 & Mersenne, 40th known \\
	4  & $2^{13,466,917}-1$ & 4,053,946 & 2001 & Mersenne, M-39 \\
	5  & $28,433 \cdot 2^{7,830,457}+1$ & 2,357,207 & 2004 & Generalized Mersenne \\
	6  & $2^{ 6,972,593}-1$ & 2,098,960 & 1999 & Mersenne, M-38 \\
	7  & $5,359 \cdot 2^{5,054,502}+1$ & 1,521,561 & 2003 & Generalized Mersenne \\
	8  & $2^{ 3,021,377}-1$ &   909,526 & 2001 & Mersenne, M-37 \\
	9  & $2^{ 2,976,221}-1$ &   895,932 & 2001 & Mersenne, M-36 \\
	10 & $1,372,930^{131,072}+1$ &   804,474 & 2003 & Generalized Fermat\footnotemark \\

\hline
\end{tabular}
\caption{The 10 largest known primes and its particular number types (as of May 2005)}
\end{center}
\end{table} 
\footnotetext{$ 1,372,930^{131,072} + 1 = 1,372,930^{(2^{17})}+1 $ } 
%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

The largest known prime is a Mersenne prime.
This prime was found by the \hyperlink{GIMPS-project}{GIMPS project}.

Within the largest known primes there are also numbers of the type
\hyperlink{generalizedMersennenumbers}{generalized Mersenne number}
(chapter~\ref{generalized-mersenne-no1})
and 
\hyperlink{generalizedFermatprimes}{generalized Fermat numbers}
(chapter~\ref{generalized-fermat}).

% --------------------------------------------------------------------------
\subsubsection{Special number types -- Mersenne numbers and Mersenne primes} 
\hypertarget{MersenneNumbers01}{}
\label{zahlentyp_mersenne}
\index{Mersenne!number}

Almost all known huge prime numbers are special candidates, called
\index{Mersenne, Marin} {\em Mersenne numbers}\footnote{%
Marin Mersenne, French priest and mathematician, Sep 08, 1588 $-$ Sep 01, 1648.
\index{Mersenne, Marin}
}
of the form $2^p -1,$ where $p$ is
a prime. Not all Mersenne numbers are prime:

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{Factorisation} (expression as a product of prime factors)\footnote{%
  Using CrypTool\index{CrypTool} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorisation of a Number}. \\
  CrypTool can factorize in a reasonable time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CrypTool. \\
  The current factorization records are listed in chapter \ref{NoteFactorisation}.
  \index{Factorisation!factorisation records}
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \href{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}
   {\texttt{http://www.isthe.com/chongo/tech/math/prime/mersenne.html}}. \\
Also see:
      \href{http://www.utm.edu/}
   {\texttt{http://www.utm.edu/}}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221;
 ~ 3,021,377; ~ 6,972,593; \\
 ~ 13,466,917;  ~ 20,996,011;  ~ 24,036,583;  ~ 25,964,951.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus
$42$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 39            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers we know that this list is complete.
The exponents until the 40th     % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \href{http://www.mersenne.org/status.htm}
   {\texttt{http://www.mersenne.org/status.htm}}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}.

                                                 }.
The $19$th number with the exponent $4,253$ was the first with at least $1,000$ digits in  decimal system
(the mathematician Samual \index{Yates, Samual} Yates coined the expression {\em
titanic} \index{Prime number!titanic} prime for this; it was discovered by
Hurwitz in 1961); the $27$th number with the exponent $44,497$ was the first
with at least $10,000$ digits in the decimal system (Yates coined the expression
\index{Prime number!gigantic}  {\em gigantic} prime for this. These names are
now long outdated).

\vskip +25 pt
\paragraph{M-37 -- January 1998}\index{Mersenne!prime number!M-37}\mbox{}

@@

\vskip +25 pt
\paragraph{M-38 -- June 1999}\index{Mersenne!prime number!M-38}\mbox{}

@@

\vskip +25 pt
\paragraph{M-39 -- December 2001}%
\index{Mersenne!prime number!M-39}\mbox{}

The 39th Mersenne prime, called M-39, $$2^{13,466,917}-1,$$ was
published at December 6, 2001 -- more exactly, the verification of this number,
found at November 14, 2001 by the Canadian student Michael Cameron, was
successfully completed. 
This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number 
$$(924947738006701322247758 \cdots 1130073855470256259071)$$
would require around 200 pages in the Financial Times.

Right now (May 2005) all prime exponents smaller than $ 13.466.917 $ have been
tested and double-checked (see home page of the GIMPS project:
{\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}): so we can
be certain, that this is really the 39th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes.

@@

Right now (as of May 2005) already 42 Mersenne primes are known.

\vskip +25 pt
\paragraph{GIMPS}\index{GIMPS}\mbox{}
\hypertarget{GIMPS-project}{}

The GIMPS project (Great Internet Mersenne Prime Search)\index{GIMPS} was founded
in 1996 by George Woltman\index{Woltman, George} to search for new largest 
Mersenne primes 
({\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}).
Further explanations about this number type can be found under
\hyperlink{MersenneNumbers02}{Mersenne numbers} and 
\hyperlink{MersenneNumbers01}{Mersenne primes}.

Right now the GIMPS project has discovered eight
            % be_2005_UPDATEN_if-new-mersenne-prime-appears
largest Mersenne primes so far, including the largest known prime number at all.

The following table contains these Mersenne record primes\footnote{%
An up-to-date version can be found in the internet at
     \href{http://www.mersenne.org/history.htm}
  {\texttt{http://www.mersenne.org/history.htm}}.
}:

            % be_2005_UPDATEN_if-new-mersenne-prime-appears
\begin{table}[h]
\begin{center}
\begin{tabular}{|cccc|}
\hline
	{\bf Definition} & {\bf Decimal Digits} & {\bf When} & {\bf Who} \\
\hline
	$2^{25,964,951}-1$ & 7,816,230 & February 18, 2005 & Dr. Martin Nowak \\
	$2^{24,036,583}-1$ & 7,235,733 & May 15, 2004      & Josh Findley     \\
	$2^{20,996,011}-1$ & 6,320,430 & November 17, 2003 & Michael Shafer   \\
	$2^{13,466,917}-1$ & 4,053,946 & November 14, 2001 & Michael Cameron  \\
	$2^{ 6,972,593}-1$ & 2,098,960 & June 1, 1999      & Nayan Hajratwala \\
	$2^{ 3,021,377}-1$ &   909,526 & January 27, 1998  & Roland Clarkson  \\
	$2^{ 2,976,221}-1$ &   895,932 & August 24, 1997   & Gordon Spence    \\
	$2^{ 1,398,269}-1$ &   420,921 & November 1996     & Joel Armengaud   \\

\hline
\end{tabular}
\caption{The largest primes found by the GIMPS project (as of May 2005)}
\end{center}
\end{table}

%be_2005: Erzwingen, dass die Abb. noch in diesem Kapitel !

Dr. Richard Crandall\index{Crandall, Richard} discovered the advanced 
transform algorithm used by the GIMPS program. George Woltman implemented
Crandall's algorithm in machine language, thereby producing a prime-search 
program of unprecedented efficiency,
and that work led to the successful GIMPS project.

On June 1st, 2003 a possible Mersenne prime was reported to the GIMPS server, 
which was checked afterwards as usual, before it was to be published. 
Unfortunately mid June the initiator and GIMPS project leader George Woltman
had to tell, that two independent verification runs proved the number 
was composite. This was the first false positive report of a client in 7 years.

Now more than 130,000 volunteers, amateurs and experts, participate in the 
GIMPS project. They connect their computers into the so called ``primenet'', 
organized by the company entropia.

% --------------------------------------------------------------------------
\vskip +25 pt
\subsubsection{Challenge of the Electronic Frontier Foundation (EFF)}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an 
unknown donator. The participants are rewarded with a total of 500,000 USD if
they find the longest prime number. In promoting this project, the unknown
donator is not looking for the quickest computer, but rather wants to draw 
people's attention to the opportunities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}

@@

Edouard Lucas\index{Lucas, Edouard} (1842-1891) held the record for the
longest prime number for over 70 years by proving that $2^{127}-1$ is prime.
No new record is likely to last that long.

% --------------------------------------------------------------------------
\subsection{Prime number tests}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

If we look for the prime factors in order to decide whether a number is prime,
then the search takes too long, if even the smallest prime factor is enormous.
Factorising numbers using systematic computational
division or using the \hyperlink{SieveEratosthenes01}{sieve of Eratosthenes} 
\index{Eratosthenes!sieve} is only feasible using current computers for
numbers with up to around $20$ digits in the decimal system.
The biggest number factorized into its 2 almost equal prime factors 
has 200 digits 
(see \hyperlink{RSA-200-chap3}{RSA-200} in chapter \ref{NoteFactorisation}).
% be_2005_UPDATEN_if-new-factorization-record-appears

@@

\hypertarget{FermatNumbers01}{}\label{FermatNumbers01}%
In the 17th century, Fermat\footnote{%
Pierre de Fermat, French mathematician, Aug 17, 1601 -- Jan 12, 1665.
\index{Fermat, Pierre}
}
\index{Fermat, Pierre} wrote to Mersenne \index{Mersenne, Marin} that he 
presumed that all numbers of the form 
$$ f(n) = 2^{2^n} + 1 $$ 
are prime for all whole numbers $ n \geq 0$ 
(\hyperlink{FermatNumbers02}{see below}, chapter~\ref{L-FermatNumbers02}).
\index{Number!Fermat}\index{Fermat!number}

@@

\vskip +25 pt
\paragraph{Pseudo prime numbers}%
\index{Prime number!pseudo prime} \index{Number!pseudo prime}%
\hypertarget{HT-Pseudoprimenumber01}{}\label{L-Pseudoprimenumber01}%
\mbox{}
\vskip +10 pt
Numbers n that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$
but are not prime are called {\em pseudo prime numbers}
(i.e. the exponent is not a prime). 
The first pseudo prime number is $$ 341 = 11 \cdot 31 .$$

\vskip +25 pt
\paragraph{Carmichael numbers}%
\index{Number!Carmichael}%
\hypertarget{HT-Carmichael-number01}{}\label{L-Carmichael-number01}%
\mbox{}
\vskip +10 pt
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called {\em Carmichael numbers}. The first of these is
$$ 561 = 3 \cdot 11 \cdot 17 .$$

\vskip +25 pt
\paragraph{Strong pseudo prime numbers}%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller, Gary L.}\index{Rabin, Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: it is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +30 pt
\subsection{Overview special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

For ``prime number formulae'' usually very special types of numbers are used.
The following enumeration contains the most common ideas for 
``prime number formulae'', and what our current knowledge is about
very big elements of the number series: Is their primality proven?
If their are compound numbers could their prime factors be determined?

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection{Mersenne numbers $f(n) = 2^n - 1$  \quad for $ n $ prime}
    \hypertarget{MersenneNumbers02}{}
    \index{Prime number!Mersenne} \index{Mersenne!prime number}
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than around 4,000,000 digits are 
    known (M-39\index{Mersenne!prime number!M-39}):
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}
\end{itemize}
            % be_2005_UPDATEN_if-new-mersenne-prime-appears

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection
   [Generalized Mersenne numbers $f(k,n) = k \cdot 2^n \pm 1$]
   {Generalized Mersenne numbers
   $f(k,n) = k \cdot 2^n \pm 1 $  for $ n $ prime and $ k $ small prime}
   \label{generalized-mersenne-no1}
   For this first generalisation of the Mersenne 
   numbers\index{Mersenne!number!generalized}
   there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can
   be performed in practice using software such as the Proths software from
   Yves Gallot\index{Gallot, Yves}
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

% --------------------------------------------------------------------------
\vskip +10 pt
\subsubsection
   [Generalized Mersenne numbers $ f(b,n) = b^n \pm 1 $ / Cunningham project]
   {Generalized Mersenne numbers 
   $ f(b,n) = b^n \pm 1 $ / The Cunningham project}
   \hypertarget{generalizedMersennenumbers}{}

@@

    The Fermat number
    $$ f(2,145,351) = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    At that time this prime p was the largest known prime generalized Mersenne 
    number\index{Mersenne!number!generalized} and the
    5th largest known prime number at all.

    This work was done using NewPGen from Paul Jobling's, PRP from 
    George Woltman's, Proth from Yves Gallot's programs\index{Gallot, Yves} and
    also the Proth-Gallot group at St. Patrick's College, Dublin.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Density and distribution of the primes}

@@

\item Is there an infinite number of Mersenne prime numbers?
    \index{Prime number!Mersenne}\index{Mersenne!prime number}
